Efficient and Effective Metric Space Indexing for Complete Record Linkage Al: Exploring approaches to probabilistic record linkage using similarity searching (metric space indexing?)

### 2017-11-13 paper.tex
Update on Overleaf.

We propose an metric indexing approach to \emph{complete} record
linkage, which ensures all record pairs that have a similarity above
a certain similarity threshold are being \textbf{compared}, while no pairs
with a similarity below the threshold are \textbf{compared}. Our approach
leads to a one-step record linkage process which combines indexing
with comparison and classification resulting in improved
effectiveness and efficiency as we experimentally evaluate on several
real-world databases.
Parameter space: good linkage bad linkage don't know which without ground truth. M Trees always high no config -> high confidence.

@@

@@

While indexing techniques facilitate the efficient linkage of very
large data-bases~\cite{Don15}, they generally achieve scalability at
the cost of loss in linkage quality, because potentially true matching
record pairs are removed in the indexing step, leading to a reduction
in recall of the final linkage result~\cite{Chr12}. A variety of
indexing techniques, as discussed in more details in the following
section, have been proposed, ranging from simple phonetic based
blocking~\cite{Chr12} and sorting of the databases~\cite{Dra12} to
locality sensitive hashing based techniques~\cite{Kim10,Steorts2014}.
Unsupervised~\cite{Kej13,Ram15} and supervised~\cite{Bil06,Mic06} 
learning techniques of optimal blocking schemes have also been
proposed.

@@

The overall process of a record linkage application, where indexing is
preformed prior to the comparison and classification steps, as
illustrated in Fig.~\ref{fig-rl-process}, adds a further practical
complexity. Indexing, comparison and classification are often conducted
using algorithms selected (and their parameters set) based on technical
and domain expertise of the record linkage user, followed by a manual
assessment of the linkage outcomes~\cite{Chr12}. If the quality of the
resulting links is not good enough for a certain application, the
linkage process needs to be repeated with different parameter settings
and possibly also alternative algorithms selected. This can result in
a time-consuming and labor intensive iterative process. A major
challenge is often the heuristic nature of indexing, where the choice
of a certain indexing technique and its parameters (including which
attributes to use in indexing) will determine the final outcome of a
linkage.
% Repeat:
% If a certain blocking approach results in a loss of recall in the
%final linkage results, the blocking needs to be redone, an often time
%consuming process.

In this paper we explore how existing indexing techniques can lead to
loss in recall, even for record pairs that are highly similar and
likely correspond to true matches. As a result, using such techniques
can lead to \emph{incomplete} record linkage results. To overcome this
problem, we advocate the use of \emph{metric space indexing} (MSI)
approaches, where it is possible to guarantee that all record pairs,
up-to a certain distance threshold, are included in the record linkage
results. MSI also prevents the expensive detailed comparison of record
pairs that have low similarities but that need to be compared because
they have been inserted into the same block with traditional indexing
techniques. Furthermore, MSI allows us to combine the indexing,
comparison and classification steps (as traditionally conducted
independently) into one single step, as shown in
Fig.~\ref{fig-rl-process}, making the overall record linkage process
both more efficient and more effective.

@@

While MSI has been used extensively in areas such as database indexing
and general nearest neighbor search, its use in record linkage so far
has been limited. 
% Peter 20171112: a pretty dangerous statement. if wrong we're stuffed.. 
% but I am not really aware of MSI approaches in RL..
Many different MSI approaches have been developed, such as various
metric tree structures~\cite{}. 
% Peter 20171112: needs same text from you guys on MSI..
In this paper, we will use the MSI techniques of M-tree and M-file (?)
and show how their use can facilitate an improvement of both
efficiency and effectiveness of the record linkage process.

@@

@@

We now briefly review relevant related work in the areas of indexing
for record linkage (for a recent survey see Christen~\cite{Chr12b}),
and metric space indexing.

\smallskip
\textbf{Indexing for Record Linkage:}
Techniques to link records across databases have been developed over
the past five decades~\cite{Fel69}, and scaling the linkage of large
databases has been an ongoing challenge as databases are continue to
grow in size and complexity. Traditional blocking techniques use a
single or a combination of attributes (known as a \emph{blocking key})
to insert records that have the same value(s) in such a blocking key
into the same block. Only records in the same block are then combined
with each other. To overcome variations and misspellings, the values
used in blocking keys are commonly phonetically encoded using functions
such as Soundex, NYSSI, or Double-Metaphone~\cite{Chr12}. These
techniques convert a string (assumed to be a name) into a code
according to how the name sounds, assigning the same code to similar
sounding names (such as `Gail' and `Gayle' or `Dickson' and `Dixon').

A different approach to indexing is the sorted neighborhood 
method~\cite{Mon96}, where the databases to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the
values from several attributes). A sliding window is then moved over
the databases and only records within the same window are being
compared. A fixed window size, $w$, leads to a number of record
pair comparison of $O(n \times w)$, where $n = |\mathbf{D}_A| +
|\mathbf{D}_b|$, is the number of records in the two databases
$\mathbf{D}_A$ and $\mathbf{D}_B$ to be linked. Techniques that
adaptively shrink or expand the window size based on the
characteristics of the sorting key values have shown to improve both
linkage efficiency and quality~\cite{Dra12}. The drawback of sorting
based

\emph{peter 20171112 until here, to do: LSH Harra, Kim10, Stoerts, drawbacks of exiting techniques}
blocking framework papadakis TKDE 2013

% describe LSH: harra (Kim10)
% Stoerts

% learning blocking keys

Drawbacks of existing techniques: errors in blocking/sorting keys,
manually selected, is learned need manual training data, or are
expensive (if unsupervised)

@@

\textbf{Traditional first - Peter?}

@@

@@

@@

The LSH-minhash approach is a fast but inexact method of performing similarity search. It has a number of tuning parameters: \textit{shingle\_size}, the \textit{number\_of\_bands} and the \textit{band\_size}. These are explained in the following algorithm description which follows a standard LSH approach. Firstly all the fields of the record are turned into strings and concatenated. Next the strings are \textit{shingled} into a set of n-grams according to the \textit{shingle\_size} parameter. Next a set of randomly, deterministically generated hash functions are applied to each of the sets of n-grams and the smallest (the minhash) of each application is added to a signature. The number of hashes used and thus the size of the signature is determined by the product of the {number\_of\_bands} and the {band\_size}.  Lastly the signature is divided into {number\_of\_bands} segments and the elements from the band are hashed again to create an key. The original record is added to a hashmap associated with this by this key. It can be shown that the similarities between signatures generated in the second step have very close Jaccard similarity to the original string and that if an arbitrary record is run through this process the records found in the map are close to the record being searched with good precision and recall. We will further explore these properties in this paper.

The MIFile \cite{amato2014mi} approach is another inexact technique for performing similarity search. Using  MI\_file,  each record is represented by the ordering of distances from a collection of reference objects. It is based on the idea that if two records are close to each other  will have similar neighbours and will thus be represented by similar of distances from the chosen reference objects. An inverted file is created mapping from...

In order to compare two objects in the dataset, we compare the two corresponding orderings of the reference objects

{\textbf{need to say something about the operations supported somewhere}}

coverage: proportion of all rec pairs with a certain similarity
(independent of their match status) that are being generated by
a blocking  technique

transitive closure is a problem if we have a blocking technique with overlapping blocks.

define blocking: only aimed at improving computational aspects by reducing the number of record pairs to be compared

@@

\section{Experiments and Results}
\label{sec-data}

@@

\emph{Peter, 31 Oct: We need to be careful not to reveal our identities
as the submission is double-blind - so no mention of DS.}
\textbf{I think we need to say something about this to provide context; there is none currently.}

% Peter, 31 Oct: can the following be shortened to 1 paragraph?

We also experiment over two datasets prepared by researchers pursuing previous projects \cite{reid2002} and \cite{reid2006}, which act as pseudo subsets of the Digitising Scotland data set. One data set contains records of vital events registered on the Isle of Skye, a rural district, while the other contains records from Kilmarnock, an industrial town. Both cover the period 1861-1901. These are different types of communities, with different family structures and name distributions \cite{reid2002}. Both datasets are encoded with the same schemata and include names, sex, addresses and the names of each person's father and mother. Ground truth is provided using demographer encoded identifiers which provide (incomplete) linkage between the records.  The Isle of Skye dataset comprises around 17,600 birth records, 12,300 death records, and 2,700 marriage records.
The demographers  identified:
\begin{itemize}
\item 4,300 family groups of siblings, with a mean family size of 3.9 siblings and a maximum family size of 16,
\item 2,900 links from a birth record to the child's death record,
\item 60 links from a marriage record to the groom's death record, and
\item 100 links from a marriage record to the bride's death record.
\end{itemize}
In the experiments described in this paper we examine links between birth and death records since these provided the highest levels of ground truth.

The Kilmarnock data set contains around 38,400 birth records, 23,700 death records, and 8,700 marriage records. Here, the demographers identified:
\begin{itemize}
\item 13,100 family groups of siblings, with a mean family size of 2.8 siblings and a  maximum family size of 16,
\item 8,300 links from a birth record to the child's death record,
\item 30 links from a marriage record to the groom's death record, and
\item 50 links from a marriage record to the bride's death record.
\item 
\end{itemize}
Again only links between birth and death have been used in the experiments in this paper.

>>>>

We propose a metric indexing approach to \emph{complete} record
linkage, which ensures all record pairs that have a similarity above
a certain similarity threshold are being \textbf{compared}, while no pairs
with a similarity below the threshold are \textbf{compared}. Our approach
leads to a one-step record linkage process which combines indexing
with comparison and classification resulting in improved
effectiveness and efficiency as we experimentally evaluate on several
real-world databases.
Parameter space: good linkage bad linkage don't know which without ground truth. M Trees always high no config -> high confidence.

@@

% 1 1/2 pages maximum!

@@

While indexing techniques facilitate the efficient linkage of very
large data-bases~\cite{Don15}, they generally achieve scalability at
the cost of loss in linkage quality, because potentially true matching
record pairs are removed in the indexing step, leading to a reduction
in recall of the final linkage result~\cite{Chr12}. A variety of
indexing techniques, as discussed in more detail in the following
section, have been proposed, ranging from simple phonetic based
blocking~\cite{Chr12} and sorting of the databases~\cite{Dra12} to
locality sensitive hashing based techniques~\cite{Kim10,Steorts2014}.
Unsupervised~\cite{Kej13,Ram15} and supervised~\cite{Bil06,Mic06} 
learning techniques of optimal blocking schemes have also been
proposed.

@@

The overall process of a record linkage application, where indexing is
preformed prior to the comparison and classification steps, as
illustrated in Fig.~\ref{fig-rl-process}, adds a further practical
complexity. Indexing, comparison and classification are often conducted
using algorithms selected (and their parameters set) based on technical
and domain expertise of the user of the record linkage system, followed by a manual
assessment (auditing) of the linkage outcomes~\cite{Chr12}. If the quality of the
resulting links is not good enough for a certain application, the
linkage process needs to be repeated with different parameter settings
and possibly also alternative algorithms selected. This can result in
a time-consuming and labor intensive iterative process. A major
challenge is often the heuristic nature of indexing, where the choice
of a certain indexing technique and its parameters (including which
attributes to use in indexing) will determine the final outcome of a
linkage.
% Repeat:
% If a certain blocking approach results in a loss of recall in the
%final linkage results, the blocking needs to be redone, an often time
%consuming process.

In this paper we explore how existing indexing techniques can lead to
loss in recall, even for record pairs that are highly similar and
likely correspond to true matches. As a result, using such techniques
can lead to \emph{incomplete} record linkage results. To overcome this
problem, we advocate the use of \emph{metric space indexing} (MSI)
approaches, where it is possible to guarantee that all record pairs,
up-to a certain distance threshold, are included in the record linkage
results. MSI also prevents the expensive detailed comparison of record
pairs that have low similarities but that need to be compared because
they have been inserted into the same block with traditional indexing
techniques. Furthermore, MSI allows the combination of the indexing,
comparison and classification steps (as traditionally conducted
independently) into one single step, as shown in
Fig.~\ref{fig-rl-process}, making the overall record linkage process
both more efficient and more effective.

% Al: but is it? be careful of last sentence in final version.

@@

While MSI has been used extensively in areas such as database indexing
and general nearest neighbor search, its use in record linkage so far
has been limited. 
% Al - So take it out - I recently found and cited a paper that uses R trees later in the paper?

% Peter 20171112: a pretty dangerous statement. if wrong we're stuffed.. 
% but I am not really aware of MSI approaches in RL..
Many different MSI approaches have been developed, such as various
metric tree structures~\cite{}. 
% Peter 20171112: needs same text from you guys on MSI..
In this paper, we will use the MSI techniques of M-tree and M-file (?)
and show how their use can facilitate an improvement of both
efficiency and effectiveness of the record linkage process.

@@

% 1 page maximum!

@@

We now briefly review relevant related work in the areas of indexing
for record linkage (for a recent survey see~\cite{Chr12b,Pap16}),
and metric space indexing~\cite{Zezula2010}.

\smallskip
\textbf{Indexing for Record Linkage:}
Techniques to link records across databases have been developed for
over five decades~\cite{Fel69,New59}, and the scalability of linking
has been an ongoing challenge as databases are continue to grow in
size and complexity. Traditional blocking~\cite{Chr12b} use a single
or a combination of attributes (known as a \emph{blocking key}) to
insert records that have the same value(s) in a blocking key into
the same block. Only records within the same block are then combined
with each other. To overcome variations and misspellings, the values
used in blocking keys can be phonetically encoded using functions
such as Soundex, NYSIIS, or Double-Metaphone~\cite{Chr12}. These
techniques convert a string (assumed to be a name) into a code
according to its pronunciation, assigning the same code to similar
sounding names (such as `Gail' and `Gayle').

A different approach to indexing is the sorted neighborhood 
method~\cite{Mon96}, where the databases to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the
values from several attributes). A sliding window is then moved over
the databases and only records within the same window are being
compared. Techniques that adaptively shrink or expand the window size based on
the characteristics of the sorting key values have shown to improve
both linkage efficiency and quality~\cite{Dra12}.

% should we mention canopy clustering? best to only briefly mention
% techniques we do not use in our comparison

Locality sensitive hashing (LSH), as originally proposed to allow
efficient nearest-neighbor search in high-dimensional
spaces~\cite{Ind98}, has been employed in record linkage as a blocking
technique where attribute values are hashed multiple times, and
blocks are created from those records that share some hash values.
\emph{HARRA} is a record linkage approach based on Min-hash and LSH
which blocks, compares, and then merges linked records in an iterative
fashion, where merged records are re-hashed to improve overall linkage
quality. Steorts et al.~\cite{Steorts2014} recently evaluated two LSH
variations, with their main conclusion being that in order to get good
results LSH methods must be tuned to the particular databases being
linked. This requires good quality ground truth data which may be
unavailable or very expensive to obtain.

Existing blocking techniques have the drawbacks that they are
heuristics and commonly require domain knowledge, such as the choice
of blocking or sorting key used. Errors and variations in the
attributes used for blocking also result in records being inserted
into a wrong block and thus true matches being missed. As a result,
existing blocking techniques can lead to \emph{incomplete} linkage,
and that many of the record pairs compared in a block turn out to be
pairs with low similarities that correspond to non-matches, resulting
in \emph{inefficient} linkage.

% Peter until here, 20171113 ************

\iffalse

% commented, as overall related works should be around 1 page max

\fi

@@

@@

% end of related work should be around end of page 3!

@@

% do we need a section on preliminaries, notation, etc?
% ideally techniques are described formally, possibly via pseudo-code
% algorithmic descriptions

@@

The LSH-minhash approach is a fast but inexact method of performing similarity search. It has a number of tuning parameters: \textit{shingle\_size}, the \textit{number\_of\_bands} and the \textit{band\_size}. These are explained in the following algorithm description which follows a standard LSH approach. Firstly all the fields of the record are turned into strings and concatenated. Next the strings are \textit{shingled} into a set of n-grams according to the \textit{shingle\_size} parameter. Next a set of randomly, deterministically generated hash functions are applied to each of the sets of n-grams and the smallest (the minhash) of each application is added to a signature. The number of hashes used and thus the size of the signature is determined by the product of the {number\_of\_bands} and the {band\_size}.  Lastly the signature is divided into {number\_of\_bands} segments and the elements from the band are hashed again to create a number of keys. The original record is added to a hashmap associated with each of the keys. It can be shown that the similarities between signatures generated in the second step have very close Jaccard similarity to the original string and that if an arbitrary record is run through this process the records found in the map are close to the record being searched with good precision and recall. We will further explore these properties in this paper. To perform a {\textit{nearest-neighbour}} lookup, the query value is hashed using the procedure described above and the union of set of of values found for all hashes in the hashmap are returned. If desired, these may then be filtered by performing distance calculations on the results.

The MIFile \cite{amato2014mi} approach is inexact technique for performing similarity search based on metric spaces. Using  MI\_file,  each record is represented by the ordering of distances from a collection of reference objects. It is based on the idea that two records close to each other will have similar neighbours and will thus be represented by similar of distances from the chosen reference objects. 
In order to create the indexes, for each value to be added to the data-structure, first the k nearest reference objects (governed by a fixed parameter) are found, each is given a score (drawn from 1,2,3... based on the reference object's position from the value). Next for each score, a  mapping is created in an inverted index mapping from the value to the score of each reference object and a representation of the data associated with the reference object and the value inserted.  In order to perform a \textit{nearest-neighbour} lookup, first the k nearest reference objects (bounded by a second fixed parameter). Next the objects closest to these reference objects are extracted from the inverted map by choosing the neighbours with the highest score. Much of the search space may be eliminated by  keeping track of the scores that  can contribute to the nearest neighbours.

M-trees support exact search operations over a set of values over a metric space (distances must support  triangle inequality - distance(i,j) + distance(j,k ) >= distance(i,k)). In addition to supporting exact search the operations supported by M-trees are richer than the other data structures described above and include \textit{range search} (all values within distance d and some query q), \textit{nearest-neighbour} (the closest value to q), and \textit{nearestN} (the N closest n objects to q). An M-tree is a balanced tree structure comprised of fixed sized nodes that can each store up to \textit{M} children. In addition to the children, each node contains a reference to an object being indexed, a pointer and the distance to its parent, and a radius (described below). The radius of an internal node called a \textit{routing-node} represents the  furthest distance to any of its children. Thus all the children may be visualised as being contained within a ball of radius \textit{r} from the parent. M-trees are grown from the roots upward and require dynamic tree rebalancing when set of children become full (of size \textit{M}).

% TODO coverage: proportion of all rec pairs with a certain similarity (independent of their match status) that are being generated by a blocking  technique - what does above refer to? al

@@

\section{Experiments and Results}
\label{sec-expt}

@@

% Peter, 31 Oct: can the following be shortened to 1 paragraph - tick - al.

We also experiment over two datasets prepared by researchers pursuing previous projects \cite{reid2002} and \cite{reid2006}, which act as pseudo subsets for the demographic linkage on which our work is primarily focused . One data set contains records of vital events registered on the Isle of Skye, a rural district, while the other contains records from Kilmarnock, an industrial town. Both cover the period 1861-1901. These are different types of communities, with different family structures and name distributions \cite{reid2002}. Both datasets are encoded with the same schemata and include names, sex, addresses and the names of each person's father and mother. Ground truth is provided using demographer encoded identifiers which provide (incomplete) linkage between the records.  The Isle of Skye dataset comprises around 17,600 birth records, 12,300 death records, and 2,700 marriage records. Relating to the experiments described below, this dataset contains 2,900 links from a birth record to the child's death record. The Kilmarnock data set contains around 38,400 birth records, 23,700 death records, and 8,700 marriage records with a similar schema to the Skye data. In this dataset demographers identified  2,900 links from a birth record to the child's death record.
